FOODIE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern

### 2015-10-19 papers/announcement/manuscript.tex
Change project name to FOODIE
FOODIE => Fortran Object-Oriented Differential-equations Integration Environment

Why:

Due to a great idea of Damian.

This change addresses the need by:

Change all references to old FOODiE acronym.

Side effects:

A lot...

\title{FOODiE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern}

@@

\begin{abstract}
The (numerical) solution of Ordinary Differential Equations (ODEs) problems is of paramount relevance, ODEs system being an ubiquitous mathematical formulation of many physical phenomena (such as those involved in fluid dynamics, chemistry, biology, evolutionary-anthropology, \dots). The present paper is the first \emph{manifesto} of FOODiE, a library aimed to numerically solve ODEs problems by means of a clear, concise and efficient \emph{abstract} interface. FOODiE, meaning Fortran Object oriented Ordinary Differential Equations integration library, has manifolds aims: to provide a set to built-in numerical schemes that are accurate, robust, validated and efficient and to allow easy application of these schemes to (almost) all ODEs problems by means of an effective Abstract Calculus Pattern. The key idea is to allow the same solver-implementation to be applied to all ODEs problems thus avoiding the re-implementation of the ODEs solver for each different ODEs problem: code re-usability is consequently maximized, FOODiE being a general robust framework. Besides, the same framework also allows rapid development of new ODEs solvers due to the high abstraction level of the library itself.

FOODiE is a modern Fortran library which main features are:
\begin{description}
  \item[Free] FOODiE is a free software;
  \item[OOP] FOODiE is based on Object Oriented Programming paradigm;
  \item[TDD] the FOODiE development follows the Test Driven Development software process;
  \item[Accurately documented] the FOODiE documentation is based on high quality, first class solutions embedding detailed (mathematical) descriptions directly inside code sources; moreover, comprehensive hyper-linked documentations is also provided;
  \item[Comphrensive] FOODiE provides a comprehensive set of built-in solvers, namely one-step, multi-steps, explicit, implicit, predictor-corrector solvers; moreover, its Abstract Calculus Pattern allows the solution of virtually all kinds of ODEs problems;
  \item[Collaborative] the development of FOODiE takes advantage of web communications, the main project being hosted on GitHub.
\end{description}

The present paper is the first announcement of FOODiE project: the current implementation is extensively discussed and its capabilities are proved by means of tests and examples.
\end{abstract}

@@

\begin{small}
\noindent
\emph{Manuscript Title:} FOODiE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern \\
\emph{Authors:} Zaghi, S., Curcic, M., Rouson, D., Beekman, I. \\
\emph{Program title:} FOODiE \\
\emph{Journal Reference:} \\
\emph{Catalogue identifier:} \\
\emph{Licensing provisions:} GNU General Public License (GPL) v3 \\
\emph{Programming language:} Fortran (standard 2008 or newer); developed and tested with GNU gfortran 5.2 or newer \\
\emph{Computer(s) for which the program has been designed:} designed for shared-memory multi-cores workstations and for hybrid distributed/shared-memory supercomputers, but any computer system with a Fortran (2008+) compiler is suited \\
\emph{Operating system(s) for which the program has been designed:} designed for POSIX architecture and tested on GNU/Linux one \\
\emph{RAM required to execute with typical data: bytes:} $[1MB,1GB]\times core$, simulation-dependent \\
\emph{Has the code been vectorised or parallelized?:} the library is not aware of the parallel back-end, it providing a high-level models, but the provided tests suite shows parallel usage by means of MPI library and OpenMP paradigm \\
\emph{Number of processors used:} tested up to 256 \\
\emph{Supplementary material:}    \\
\emph{Keywords:} ODE, PDE, OOP, ACP, Fortran \\
\emph{CPC Library Classification:} 4.3 Differential Equations, 4.10 Interpolation, 12 Gases and Fluids \\
\emph{External routines/libraries used:} \\
\emph{CPC Program Library subprograms used:} \\
\emph{Nature of problem:} \\
Numerical integration of (general) Ordinary Differential Equations system \\
\emph{Solution method:} \\
\emph{Restrictions:} \\
\emph{Unusual features:} \\
\emph{Additional comments:} \\
\emph{Running time:} \\
\emph{References:} \\
% \begin{thebibliography}{0}
% \end{thebibliography}
\end{small}

>>>>

\title{FOODIE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern}

@@

\begin{abstract}
The (numerical) solution of Ordinary Differential Equations (ODEs) problems is of paramount relevance, ODEs system being an ubiquitous mathematical formulation of many physical phenomena (such as those involved in fluid dynamics, chemistry, biology, evolutionary-anthropology, \dots). The present paper is the first \emph{manifesto} of FOODIE, a library aimed to numerically solve ODEs problems by means of a clear, concise and efficient \emph{abstract} interface. FOODIE, meaning Fortran Object oriented Ordinary Differential Equations integration library, has manifolds aims: to provide a set to built-in numerical schemes that are accurate, robust, validated and efficient and to allow easy application of these schemes to (almost) all ODEs problems by means of an effective Abstract Calculus Pattern. The key idea is to allow the same solver-implementation to be applied to all ODEs problems thus avoiding the re-implementation of the ODEs solver for each different ODEs problem: code re-usability is consequently maximized, FOODIE being a general robust framework. Besides, the same framework also allows rapid development of new ODEs solvers due to the high abstraction level of the library itself.

FOODIE is a modern Fortran library which main features are:
\begin{description}
  \item[Free] FOODIE is a free software;
  \item[OOP] FOODIE is based on Object Oriented Programming paradigm;
  \item[TDD] the FOODIE development follows the Test Driven Development software process;
  \item[Accurately documented] the FOODIE documentation is based on high quality, first class solutions embedding detailed (mathematical) descriptions directly inside code sources; moreover, comprehensive hyper-linked documentations is also provided;
  \item[Comphrensive] FOODIE provides a comprehensive set of built-in solvers, namely one-step, multi-steps, explicit, implicit, predictor-corrector solvers; moreover, its Abstract Calculus Pattern allows the solution of virtually all kinds of ODEs problems;
  \item[Collaborative] the development of FOODIE takes advantage of web communications, the main project being hosted on GitHub.
\end{description}

The present paper is the first announcement of FOODIE project: the current implementation is extensively discussed and its capabilities are proved by means of tests and examples.
\end{abstract}

@@

\begin{small}
\noindent
\emph{Manuscript Title:} FOODIE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern \\
\emph{Authors:} Zaghi, S., Curcic, M., Rouson, D., Beekman, I. \\
\emph{Program title:} FOODIE \\
\emph{Journal Reference:} \\
\emph{Catalogue identifier:} \\
\emph{Licensing provisions:} GNU General Public License (GPL) v3 \\
\emph{Programming language:} Fortran (standard 2008 or newer); developed and tested with GNU gfortran 5.2 or newer \\
\emph{Computer(s) for which the program has been designed:} designed for shared-memory multi-cores workstations and for hybrid distributed/shared-memory supercomputers, but any computer system with a Fortran (2008+) compiler is suited \\
\emph{Operating system(s) for which the program has been designed:} designed for POSIX architecture and tested on GNU/Linux one \\
\emph{RAM required to execute with typical data: bytes:} $[1MB,1GB]\times core$, simulation-dependent \\
\emph{Has the code been vectorised or parallelized?:} the library is not aware of the parallel back-end, it providing a high-level models, but the provided tests suite shows parallel usage by means of MPI library and OpenMP paradigm \\
\emph{Number of processors used:} tested up to 256 \\
\emph{Supplementary material:}    \\
\emph{Keywords:} ODE, PDE, OOP, ACP, Fortran \\
\emph{CPC Library Classification:} 4.3 Differential Equations, 4.10 Interpolation, 12 Gases and Fluids \\
\emph{External routines/libraries used:} \\
\emph{CPC Program Library subprograms used:} \\
\emph{Nature of problem:} \\
Numerical integration of (general) Ordinary Differential Equations system \\
\emph{Solution method:} \\
\emph{Restrictions:} \\
\emph{Unusual features:} \\
\emph{Additional comments:} \\
\emph{Running time:} \\
\emph{References:} \\
% \begin{thebibliography}{0}
% \end{thebibliography}
\end{small}

### 2015-10-19 papers/announcement/src/API/API.tex
Change project name to FOODIE
FOODIE => Fortran Object-Oriented Differential-equations Integration Environment

Why:

Due to a great idea of Damian.

This change addresses the need by:

Change all references to old FOODiE acronym.

Side effects:

A lot...

In this section we review the FOODiE API providing a detailed discussion of the implementation choices.

As aforementioned, the programming language used is the Fortran 2008 standard, that is a minor revision of the previous Fortran 2003 standard. Such a new Fortran idioms provide (among other useful features) an almost complete support for OOP, in particular for ADT concept. Fortran 2003 has introduced the \emph{abstract derived type}: it is a derived type suitable to serve as \emph{contract} for concrete type-extensions that has not any actual implementations, rather it provides a well-defined set of type bound procedures interfaces, that in Fortran nomenclature are called \emph{deferred} procedures. Using such an abstract definition, we can implement algorithms operating on only this abstract type and on \emph{all its concrete extensions}. This is the key feature of FOODiE library: all the above described ODE solvers are implemented on the knowledge of \emph{only one abstract type}, allowing an implementation-style based on a very high-level syntax. In the meanwhile, client codes must implement their own IVPs extending only one simple abstract type.

In the subsection \ref{subsec:type_integrand} a review of the FOODiE main ADT, the \emph{integrand} type, is provided, while subsections \ref{subsec:solver_euler}, \ref{subsec:solver_tvd_rk}, \ref{subsec:solver_ls_rk}, \ref{subsec:solver_ab} and \ref{subsec:solver_leapfrog} cover the API of the currently implemented solvers.

It is worth noting that all FOODiE public \emph{entities} (ADT and solvers) must be accessed by the FOODiE module, see listing \ref{list:main_module} for an example on how access to all public FOODiE entities.

\begin{lstlisting}[firstnumber=1,style=code,caption={usage example importing all public entities of FOODiE main module},label={list:main_module}]
use foodie, only: integrand, &
                  adams_bashforth_integrator, &
                  adams_moulton_integrator, &
                  adams_bashforth_moulton_integrator, &
                  euler_explicit_integrator, &
                  leapfrog_integrator, &
                  ls_runge_kutta_integrator, &
                  tvd_runge_kutta_integrator
! or simply
use foodie
\end{lstlisting}

\subsection{The main FOODiE Abstract Data Type: the integrand type}\label{subsec:type_integrand}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={integrand type definition},label={list:type_integrand}]
type, abstract :: integrand
  !< Abstract type for building FOODiE ODE integrators.
  contains
    ! public deferred procedures that concrete integrand-field must implement
    procedure(time_derivative),       pass(self), deferred, public:: t
    ! operators
    procedure(symmetric_operator),   pass(lhs), deferred, public:: integrand_multiply_integrand
    procedure(integrand_op_real),    pass(lhs), deferred, public:: integrand_multiply_real
    procedure(real_op_integrand),    pass(rhs), deferred, public:: real_multiply_integrand
    procedure(symmetric_operator),   pass(lhs), deferred, public:: add
    procedure(symmetric_operator),   pass(lhs), deferred, public:: sub
    procedure(assignment_integrand), pass(lhs), deferred, public:: assign_integrand
    ! operators overloading
    generic, public:: operator(+) => add
    generic, public:: operator(-) => sub
    generic, public:: operator(*) => integrand_multiply_integrand, &
                                     real_multiply_integrand, &
                                     integrand_multiply_real
    generic, public:: assignment(=) => assign_integrand
endtype integrand
\end{lstlisting}

The \emph{integrand} type does not implement any actual integrand field, it being and abstract type. It only specifies which deferred procedures are necessary for implementing an actual concrete integrand type that can use a FOODiE solver.

@@

Commonly, into the concrete implementation of this deferred abstract procedure clients embed the actual ODE equations being solved. As an example, for the Burgers equation, that is a Partial Differential Equations (PDE) system involving also a boundary value problem, this procedure embeds the spatial operator that convert the PDE to a system of algebraic ODE. As a consequence, the eventual concrete implementation of this procedure can be very complex and errors-prone. Nevertheless, the FOODiE solvers are implemented only on the above abstract interface, thus emancipating the solvers implementation from any concrete complexity.

@@

This interface defines a class of procedures operating on 2 \emph{integrand} objects, namely it is used for the definition of the operators \emph{multiplication}, \emph{summation} and \emph{subtraction} of integrand objects. These operators are used into the above described ODE solvers, for example see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of such a procedures class are strictly dependent on the concrete extension of the integrand type. From the FOODiE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the operator, and the second argument, the right-hand-side of the operator, are two integrand object and the returned object is still an integrand one.

@@

These two interfaces are necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the multiplication of an integrand object for a real number, circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of these procedures are strictly dependent on the concrete extension of the integrand type. From the FOODiE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the operator, and the second argument, the right-hand-side of the operator, are an integrand object and real number of viceversa and the returned object is still an integrand one.

@@

The assignment statement is necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the assignment of an integrand object by another one, circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of this assignment is strictly dependent on the concrete extension of the integrand type. From the FOODiE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the assignment, and the second argument, the right-hand-side of the assignment, are two integrand objects.

@@

The explicit forward Euler solver is exposed (by the FOODiE main module that must imported, see listing \ref{list:main_module}) as a single derived type (that is a standard convention for all FOODiE solvers) named \emph{euler\_explicit\_integrator}. It provides the type bound procedure (also referred as \emph{method}) \emph{integrate} for integrating in time an \emph{integrand} object, or any of its polymorphic concrete extensions. Consequently, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_euler_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit forward Euler integrator},label={list:define_euler_solver}]
use FOODiE, only: euler_explicit_integrator
type(euler_explicit_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_tvd_rk_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit TVD/SSP Runge-Kutta integrator},label={list:define_tvd_rk_solver}]
use FOODiE, only: tvd_runge_kutta_integrator
type(tvd_runge_kutta_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_ls_rk_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit low storage Runge-Kutta integrator},label={list:define_ls_rk_solver}]
use FOODiE, only: ls_runge_kutta_integrator
type(ls_runge_kutta_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_ab_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit Adams-Bashforth integrator},label={list:define_ab_solver}]
use FOODiE, only: adams_bashforth_integrator
type(adams_bashforth_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_am_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an implicit Adams-Moulton integrator},label={list:define_am_solver}]
use FOODiE, only: adams_moulton_integrator
type(adams_moulton_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_lf_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit Leapfrog integrator},label={list:define_lf_solver}]
use FOODiE, only: leapfrog_integrator
type(leapfrog_integrator) :: integrator
\end{lstlisting}

@@

Table \ref{tab:foodie-vs-math} presents a comparison of the relevant parts of equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:RK-stage}, \ref{eq:RK-ls}, \ref{eq:AB} and \ref{eq:leapfrog} with the corresponding FOODiE implementations reported in listings \ref{list:euler_integrate}, \ref{list:tvd_rk_integrate}, \ref{list:ls_rk_integrate}, \ref{list:ab_integrate} and \ref{list:lf_integrate}, respectively. This comparison proves that the \emph{integrand} ADT has allowed a very high-level implementation syntax. The Fortran implementation is almost equivalent to the rigorous mathematical formulation. This aspect directly implies that the implementation of a ODE solver into the FOODiE library is very clear, concise and less-errors-prone than an \emph{hard-coded} implementation where the solvers must be implemented for each specific definition of the integrand type, it being not abstract.

\begin{table}[!ht]
  \centering
  \caption{Comparison between rigorous mathematical formulation and FOODiE high-level implementation; the syntax "$(s)$" and "$(ss)$" imply the summation operation}\label{tab:foodie-vs-math}
  \resizebox{0.99\textwidth}{!}{%
  \begin{tabular}{c|cc}
    \toprule
    {\sc Solver} & {\sc Mathematical formulation} & {\sc FOODiE implementation} \\
    \hline
    explicit forward Euler & $U\left(t^{n+1}\right) = U\left(t^n\right) +\Delta t \cdot R\left[t^n, U\left(t^n\right)\right]$ & $U = U + U\%t(t=t) * Dt$\\
    \hline
    TVD/SSP Runge-Kutta & $\begin{matrix}
  K_s = R\left(t^n + \gamma_s \Delta t, U^n+\Delta t\sum\limits_{l=1}^{s-1}{\alpha_{s,l} K_l} \right) \\
  U^{n+1} = U^{n} + \Delta t \cdot\sum\limits_{s=1}^{N_{s}}{\beta_s K_s}
\end{matrix}$ & $\begin{matrix}
  stage(s) = stage(s) + stage(ss) * (Dt * self\%alph(s, ss)) \\
  U = U +  stage(s) * (Dt * self\%beta(s))
\end{matrix}$\\
    \hline
    low storage Runge-Kutta & $\begin{matrix}
  K_2 = A_s K_2 + \Delta t \cdot R\left(t^n + C_s \Delta t, K_1\right) \\
  K_1 = K_1 + B_s K_2
\end{matrix}$ & $\begin{matrix}
  stage(2) = stage(2) * self\%A(s) + \\ + stage(1)\%t(t=t + self\%C(s) * Dt) * Dt \\
  stage(1) = stage(1) + stage(2) * self\%B(s)
\end{matrix}$\\
    \hline
  explicit Adams-Bashforth & $\begin{matrix}U\left(t^{n+N_s}\right) = U\left(t^{n+N_s-1}\right) + \\ +\Delta t \sum_{s=1}^{n+N_s}{ b_s \cdot R\left[t^{n+s-1}, U\left(t^{n+s-1}\right)\right]}\end{matrix}$ & $U = U + U\%t(n=s, t=t(s)) * (Dt * self\%b(s))$\\
    \hline
    explicit Leapfrog & $ U\left(t^{n+2}\right) = U\left(t^{n}\right) + 2\Delta t \cdot R\left[t^{n+1}, U\left(t^{n+1}\right)\right]$ & $U = U\%previous\_step(n=1) + U\%t(n=2, t=t) * (Dt * 2.)$\\

>>>>

In this section we review the FOODIE API providing a detailed discussion of the implementation choices.

As aforementioned, the programming language used is the Fortran 2008 standard, that is a minor revision of the previous Fortran 2003 standard. Such a new Fortran idioms provide (among other useful features) an almost complete support for OOP, in particular for ADT concept. Fortran 2003 has introduced the \emph{abstract derived type}: it is a derived type suitable to serve as \emph{contract} for concrete type-extensions that has not any actual implementations, rather it provides a well-defined set of type bound procedures interfaces, that in Fortran nomenclature are called \emph{deferred} procedures. Using such an abstract definition, we can implement algorithms operating on only this abstract type and on \emph{all its concrete extensions}. This is the key feature of FOODIE library: all the above described ODE solvers are implemented on the knowledge of \emph{only one abstract type}, allowing an implementation-style based on a very high-level syntax. In the meanwhile, client codes must implement their own IVPs extending only one simple abstract type.

In the subsection \ref{subsec:type_integrand} a review of the FOODIE main ADT, the \emph{integrand} type, is provided, while subsections \ref{subsec:solver_euler}, \ref{subsec:solver_tvd_rk}, \ref{subsec:solver_ls_rk}, \ref{subsec:solver_ab} and \ref{subsec:solver_leapfrog} cover the API of the currently implemented solvers.

It is worth noting that all FOODIE public \emph{entities} (ADT and solvers) must be accessed by the FOODIE module, see listing \ref{list:main_module} for an example on how access to all public FOODIE entities.

\begin{lstlisting}[firstnumber=1,style=code,caption={usage example importing all public entities of FOODIE main module},label={list:main_module}]
use foodie, only: integrand, &
                  adams_bashforth_integrator, &
                  adams_moulton_integrator, &
                  adams_bashforth_moulton_integrator, &
                  euler_explicit_integrator, &
                  leapfrog_integrator, &
                  ls_runge_kutta_integrator, &
                  tvd_runge_kutta_integrator
! or simply
use foodie
\end{lstlisting}

\subsection{The main FOODIE Abstract Data Type: the integrand type}\label{subsec:type_integrand}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={integrand type definition},label={list:type_integrand}]
type, abstract :: integrand
  !< Abstract type for building FOODIE ODE integrators.
  contains
    ! public deferred procedures that concrete integrand-field must implement
    procedure(time_derivative),       pass(self), deferred, public:: t
    ! operators
    procedure(symmetric_operator),   pass(lhs), deferred, public:: integrand_multiply_integrand
    procedure(integrand_op_real),    pass(lhs), deferred, public:: integrand_multiply_real
    procedure(real_op_integrand),    pass(rhs), deferred, public:: real_multiply_integrand
    procedure(symmetric_operator),   pass(lhs), deferred, public:: add
    procedure(symmetric_operator),   pass(lhs), deferred, public:: sub
    procedure(assignment_integrand), pass(lhs), deferred, public:: assign_integrand
    ! operators overloading
    generic, public:: operator(+) => add
    generic, public:: operator(-) => sub
    generic, public:: operator(*) => integrand_multiply_integrand, &
                                     real_multiply_integrand, &
                                     integrand_multiply_real
    generic, public:: assignment(=) => assign_integrand
endtype integrand
\end{lstlisting}

The \emph{integrand} type does not implement any actual integrand field, it being and abstract type. It only specifies which deferred procedures are necessary for implementing an actual concrete integrand type that can use a FOODIE solver.

@@

Commonly, into the concrete implementation of this deferred abstract procedure clients embed the actual ODE equations being solved. As an example, for the Burgers equation, that is a Partial Differential Equations (PDE) system involving also a boundary value problem, this procedure embeds the spatial operator that convert the PDE to a system of algebraic ODE. As a consequence, the eventual concrete implementation of this procedure can be very complex and errors-prone. Nevertheless, the FOODIE solvers are implemented only on the above abstract interface, thus emancipating the solvers implementation from any concrete complexity.

@@

This interface defines a class of procedures operating on 2 \emph{integrand} objects, namely it is used for the definition of the operators \emph{multiplication}, \emph{summation} and \emph{subtraction} of integrand objects. These operators are used into the above described ODE solvers, for example see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of such a procedures class are strictly dependent on the concrete extension of the integrand type. From the FOODIE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the operator, and the second argument, the right-hand-side of the operator, are two integrand object and the returned object is still an integrand one.

@@

These two interfaces are necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the multiplication of an integrand object for a real number, circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of these procedures are strictly dependent on the concrete extension of the integrand type. From the FOODIE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the operator, and the second argument, the right-hand-side of the operator, are an integrand object and real number of viceversa and the returned object is still an integrand one.

@@

The assignment statement is necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the assignment of an integrand object by another one, circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of this assignment is strictly dependent on the concrete extension of the integrand type. From the FOODIE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the assignment, and the second argument, the right-hand-side of the assignment, are two integrand objects.

@@

The explicit forward Euler solver is exposed (by the FOODIE main module that must imported, see listing \ref{list:main_module}) as a single derived type (that is a standard convention for all FOODIE solvers) named \emph{euler\_explicit\_integrator}. It provides the type bound procedure (also referred as \emph{method}) \emph{integrate} for integrating in time an \emph{integrand} object, or any of its polymorphic concrete extensions. Consequently, for using such a solver it must be previously defined as an instance of the exposed FOODIE integrator type, see listing \ref{list:define_euler_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit forward Euler integrator},label={list:define_euler_solver}]
use FOODIE, only: euler_explicit_integrator
type(euler_explicit_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODIE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODIE integrator type, see listing \ref{list:define_tvd_rk_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit TVD/SSP Runge-Kutta integrator},label={list:define_tvd_rk_solver}]
use FOODIE, only: tvd_runge_kutta_integrator
type(tvd_runge_kutta_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODIE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODIE integrator type, see listing \ref{list:define_ls_rk_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit low storage Runge-Kutta integrator},label={list:define_ls_rk_solver}]
use FOODIE, only: ls_runge_kutta_integrator
type(ls_runge_kutta_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODIE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODIE integrator type, see listing \ref{list:define_ab_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit Adams-Bashforth integrator},label={list:define_ab_solver}]
use FOODIE, only: adams_bashforth_integrator
type(adams_bashforth_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODIE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODIE integrator type, see listing \ref{list:define_am_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an implicit Adams-Moulton integrator},label={list:define_am_solver}]
use FOODIE, only: adams_moulton_integrator
type(adams_moulton_integrator) :: integrator
\end{lstlisting}

@@

As common for FOODIE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODIE integrator type, see listing \ref{list:define_lf_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit Leapfrog integrator},label={list:define_lf_solver}]
use FOODIE, only: leapfrog_integrator
type(leapfrog_integrator) :: integrator
\end{lstlisting}

@@

Table \ref{tab:foodie-vs-math} presents a comparison of the relevant parts of equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:RK-stage}, \ref{eq:RK-ls}, \ref{eq:AB} and \ref{eq:leapfrog} with the corresponding FOODIE implementations reported in listings \ref{list:euler_integrate}, \ref{list:tvd_rk_integrate}, \ref{list:ls_rk_integrate}, \ref{list:ab_integrate} and \ref{list:lf_integrate}, respectively. This comparison proves that the \emph{integrand} ADT has allowed a very high-level implementation syntax. The Fortran implementation is almost equivalent to the rigorous mathematical formulation. This aspect directly implies that the implementation of a ODE solver into the FOODIE library is very clear, concise and less-errors-prone than an \emph{hard-coded} implementation where the solvers must be implemented for each specific definition of the integrand type, it being not abstract.

\begin{table}[!ht]
  \centering
  \caption{Comparison between rigorous mathematical formulation and FOODIE high-level implementation; the syntax "$(s)$" and "$(ss)$" imply the summation operation}\label{tab:foodie-vs-math}
  \resizebox{0.99\textwidth}{!}{%
  \begin{tabular}{c|cc}
    \toprule
    {\sc Solver} & {\sc Mathematical formulation} & {\sc FOODIE implementation} \\
    \hline
    explicit forward Euler & $U\left(t^{n+1}\right) = U\left(t^n\right) +\Delta t \cdot R\left[t^n, U\left(t^n\right)\right]$ & $U = U + U\%t(t=t) * Dt$\\
    \hline
    TVD/SSP Runge-Kutta & $\begin{matrix}
  K_s = R\left(t^n + \gamma_s \Delta t, U^n+\Delta t\sum\limits_{l=1}^{s-1}{\alpha_{s,l} K_l} \right) \\
  U^{n+1} = U^{n} + \Delta t \cdot\sum\limits_{s=1}^{N_{s}}{\beta_s K_s}
\end{matrix}$ & $\begin{matrix}
  stage(s) = stage(s) + stage(ss) * (Dt * self\%alph(s, ss)) \\
  U = U +  stage(s) * (Dt * self\%beta(s))
\end{matrix}$\\
    \hline
    low storage Runge-Kutta & $\begin{matrix}
  K_2 = A_s K_2 + \Delta t \cdot R\left(t^n + C_s \Delta t, K_1\right) \\
  K_1 = K_1 + B_s K_2
\end{matrix}$ & $\begin{matrix}
  stage(2) = stage(2) * self\%A(s) + \\ + stage(1)\%t(t=t + self\%C(s) * Dt) * Dt \\
  stage(1) = stage(1) + stage(2) * self\%B(s)
\end{matrix}$\\
    \hline
  explicit Adams-Bashforth & $\begin{matrix}U\left(t^{n+N_s}\right) = U\left(t^{n+N_s-1}\right) + \\ +\Delta t \sum_{s=1}^{n+N_s}{ b_s \cdot R\left[t^{n+s-1}, U\left(t^{n+s-1}\right)\right]}\end{matrix}$ & $U = U + U\%t(n=s, t=t(s)) * (Dt * self\%b(s))$\\
    \hline
    explicit Leapfrog & $ U\left(t^{n+2}\right) = U\left(t^{n}\right) + 2\Delta t \cdot R\left[t^{n+1}, U\left(t^{n+1}\right)\right]$ & $U = U\%previous\_step(n=1) + U\%t(n=2, t=t) * (Dt * 2.)$\\
